\hydrogen{} and \carbon{} Network Strategies for the European Energy System

### 2023-12-21 report/submissions/journal/paper/paper.tex
prefinal internal reivew

\title{Evaluating Hydrogen and Carbon Network Strategies for the European Energy System}
% \title{Competition and Synergies of \hydrogen{} and \carbon{} Networks in Europe}

@@

\begin{abstract}
    Hydrogen and carbon dioxide transport can both play an important role in climate-neutral energy systems. Hydrogen networks help serve regions with high energy demand, while excess emissions are transported away in carbon dioxide networks. Regarding the synthesis of carbonaceous fuels, it is less clear which input should be transported: hydrogen to carbon point sources or carbon to low-cost hydrogen. We explore the potential synergies and competition of both networks in a cost-optimal carbon-neutral European energy system. In direct comparison, a hydrogen network is more cost-effective than a carbon network, since it serves to transport hydrogen to demand and to point source of carbon for utilization. However, in a hybrid scenario where both networks are present, the carbon network effectively complements the hydrogen network, promoting carbon capture from biomass and reducing reliance on direct air capture. This study shows how the European energy system in future net-zero and net-negative carbon dioxide emissions scenarios benefits from integrating both networks.
\end{abstract}

@@

The transition to a climate-neutral European economy is a pressing challenge that demands coordinated action across various energy sectors. While management of both carbon dioxide (\carbon{}) and hydrogen (\hydrogen{}) is considered a critical component of this transition, a gap exists in understanding how new hydrogen infrastructure effectively interacts with carbon management technologies, including carbon capture, transport, storage, utilization, and sequestration. Hydrogen is considered in several sectors in industry not amenable to electrification, and offers an efficient way to transport and store energy over long distances and periods, and can be produced from electrolysis or steam methane reforming (SMR), for example. Captured carbon can be stored in geological formations to permanently extract it from the atmosphere, a process known as carbon sequestration (CS). Additionally, carbon can be combined with hydrogen to produce carbonaceous fuels to supply for example synthetic kerosene for aviation, synthetic methanol for shipping or synthetic methane for industrial feedstock. From hereon, we jointly refer to the production of these carbonaceous fuels as Carbon Utilization (CU). Carbon can be effectively captured from industrial processes and the combustion of biomass, fossil fuels, or synthetic carbonaceous fuels through point-source carbon capture (CC) techniques, or harvested from the atmosphere using direct air capture (DAC). Both \carbon{} and \hydrogen{} networks are likely to play a crucial role in the cost-effective integration of these technologies to enable net-zero economies.

Recently, policymakers and industry in Europe have started developing hydrogen and carbon management strategies~\cite{GermanyDevelopingStrategy2023,CarbonManagementStrategie}, planning infrastructure components~\cite{CONetz}, and committing to the first carbon utilization projects~\cite{EFuelsPilotPlant2022,OrstedAssumesFull,GROUNDBREAKINGEFUELPRODUCTION,DLREfuelsDLR}. With the European Union's goal of achieving climate neutrality by 2050, a wide range of programs, funding models and initiatives have been established with dedicated instruments supporting both the ramp-up of the hydrogen and carbon economy~\cite{eu2023netzero,europeangreendeal,europeaninnovationfund}.
Collaborative industry initiatives like the European Hydrogen Backbone~\cite{gasforclimateEuropeanHydrogenBackbone2022} or the Hydrogen Infrastructure Map~\cite{H2InfrastructureMap} showcase the potential of hydrogen as a fuel and energy carrier. Some natural gas pipelines have already been repurposed to transport hydrogen~\cite{RohrFreiFuer}. Business models from companies like Tree Energy Solutions~\cite{TESHydrogenLife2023}, Carbfix~\cite{WeTurnCO2}, and Equinor~\cite{adomaitisEquinorRWEBuild2023} advertise carbon management hubs that provide green hydrogen, methane, and synthetic fuels on the one hand and offer to purchase \carbon{} on the other hand. The Northern Lights project in Norway~\cite{NorthernLightsWhat} is planning with a transport and sequestration capacity of 1.5~Mt \carbon{} per year to be operative in 2024, expanding to a targeted scale of 5~Mt per year of sequestration by 2030.
The market potential for carbon capture and sequestration in Europe is underscored by the Capture Map's estimated potential~\cite{ToolsGreenTransition} of 1.7 Gt carbon annually from point sources, accounting for about half of the continent's emissions, as well as estimated sequestration capacity potentials of up to 3~Gt/a~\cite{europeancommissionEuropeanCO2Storage}. However, it's important to note that carbon sequestration technology, unlike carbon capture, is not yet fully mature and its full potential and implications remain to be explored in depth.
To advance the carbon economy, the Clean Air Task Force emphasizes the importance of developing a carbon transport system in Europe alongside a hydrogen network~\cite{lockwoodEuropeanStrategyCarbon}. Carbon pipelines, considered a mature technology, have seen widespread installations in the United States and Canada, primarily to supply \carbon{} to enhanced oil recovery~\cite{righettiSitingCarbonDioxide2017,friedmannNETZEROGEOSPHERICRETURN}.

Up to this point, it remains unclear how the two transport systems of hydrogen and carbon may complement or even compete with each other. Both can bridge the misalignment of sources and sinks for carbon and hydrogen. A hydrogen network can supply regions with geographically fixed hydrogen demand, such as for steel production, with hydrogen from regions with the best renewable resources as well as enable CU at the site of point-source CC. On the other hand, the carbon transport system allows for transporting captured carbon to regions with sequestration potentials or high-quality renewable resources, the latter enabling cost-effective CU.

In the literature, the two network approaches and underlying technologies have been discussed in several of publications, all of which, however, dealt with the isolated aspects~\cite{bakkenLinearModelsOptimization2008,morbeeOptimisedDeploymentEuropean2012,stewartFeasibilityEuropeanwideIntegrated2014,oeiModelingCarbonCapture2014,elahiMultiperiodLeastCost2014,burandtDecarbonizingChinaEnergy2019,middletonSimCCSOpensourceTool2020,bjerketvedtOptimalDesignCost2020,weiProposedGlobalLayout2021,damoreOptimalDesignEuropean2021,becattiniCarbonDioxideCapture2022,neumannBenefitsHydrogenNetwork2022}. Such techno-economic models, in contrast for example to Integrated Assessment Models, can account for the spatial distribution of carbon sources and sinks which are crucial to provide a holistic view of the energy system and its technological interactions. Neumann et al.~\cite{neumannBenefitsHydrogenNetwork2022} demonstrated the interchangeability of hydrogen and electricity grid expansion in a climate-neutral European energy system at the expense of higher investments. The underlying, highly-resolved model encompasses most relevant sectors, however it neglects the option to transport carbon across regions.
Morbee et al.~\cite{morbeeOptimisedDeploymentEuropean2012} optimizes the topology and capacity of a \carbon{} network in Europe, but only considers the power sector without co-optimizing renewable deployment. This limited sectoral scope cannot capture important dynamics of carbon management, since it neglects the sectors like industry that will need to handle most \carbon{} in the future.
A comprehensive example is found in~\cite{becattiniCarbonDioxideCapture2022}, which presents a mixed-integer model to optimize the time-evolution of a \carbon{} transport system in Switzerland, connecting to a remote sequestration site in Norway. However, this limited spatial scope fails to consider other sequestration sites and co-benefits from connecting the \carbon{} network to neighboring countries.
Other high-resolution energy system models do not feature detailed CU technologies and \carbon{} transport. Euro-Calliope, for example, does neither feature carbon nor hydrogen transport, and thus forces generic synthetic fuel production to rely on captive electrolysis and DAC in the same region~\cite{pickeringDiversityOptionsEliminate2022}.

No study has yet considered the co-optimization and comprehensive assessment of both \carbon{} and \hydrogen{} networks in a fully sector-coupled energy system. However, such an assessment is strongly needed to identify realms of competition and synergy between hydrogen and carbon management technologies. In this paper, we present a detailed study of the European energy system for 2050, which includes high geographical resolution and a comprehensive representation of carbon management technologies. The study is conducted using the PyPSA-Eur energy system model which optimizes operation and investments in all relevant energy sectors to supply projected energy demands for 2050 (see Fig.~\ref{fig:total-demand-bar}). It is assumed that Europe is self-sufficient in energy and does not import any fuels. This drives carbonaceous fuels production to be located at sites where both \carbon{} and \hydrogen{} can be provided at low cost.

\begin{figure}[h!]
    \includegraphics[width=\linewidth]{baseline/figures/90_nodes/demand_bar.png}
    \caption{Assumptions on exogenous demand, derived from~\cite{piamanzGeoreferencedIndustrialSites2018,muehlenpfordtTimeSeries2019,mantzosJRCIDEES20152018,NationalEmissionsReported2023,EurostatCompleteEnergyBalance,uwekrienDemandlib2023}. The figure shows the total annual energy demand for each energy source which determine the model's endogenous investments and operation. Endogenous processes can lead to higher total production volumes of some energy carriers, e.g., the demand for methanol requires more hydrogen and carbon as secondary (energy) inputs, which are not considered here. In the model, demands are defined per region and time stamp.}
    % TODO: adjust labels to show that these are exogenous assumptions
    \label{fig:total-demand-bar}
\end{figure}

@@

To investigate the competing transport dynamics between \carbon{} and \hydrogen{} networks, we contrast the following four model scenarios:
%
\begin{itemize}
    \item[] \textit{Baseline:} Neither \carbon{} nor \hydrogen{} networks are constructed.
    \item[] \textit{\carbongrid{}:} Only the \carbon{} network is developed.
    \item[] \textit{\hydrogengrid{}:} Only the \hydrogen{} network is developed.
    \item[] \textit{Hybrid:} Both \carbon{} and \hydrogen{} networks are developed.
\end{itemize}

We analyze how the deployment of carbon and hydrogen networks affects different decarbonization strategies in the models and scenarios and how investments in none or only one of the transport technologies lead to inefficient market layouts. We do this by focusing first on a net-zero \carbon{} emissions target, while limiting carbon sequestration to a total 200~Mt/a. We give a detailed description of the different technological and geographical impacts of the transportation systems. Subsequently, our analysis demonstrates the extent to which different deployment strategies remain robust under a tightened net \carbon{} removal target.

@@

All scenarios share key energy system characteristics but differ in system costs and technology use (Fig.~\ref{fig:cost_bar}). Costs in the \baselinescenario{} are the highest at 779~bn€/a. Models with \carbon{} and \hydrogen{} transport options are more cost-effective, reducing costs by 3.1\% in the \carbongrid{}, 4.8\% in \hydrogengrid{}, and 5.2\% in \hybridscenario{}.

\begin{figure}[ht!]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/cost_bar.png}
    \caption[short]{Total annual system cost subdivided into groups of technologies for the different models of the European energy system with a net-zero emission target. While in the \baselinescenario{}, the model has neither a \carbon{} nor an \hydrogen{} network, it can expand both in the \hydrogenscenario{}. ``Gas Infrastructure'' combines gas facilities for power and heat production, ``\carbon{} Infrastructure'' and ``\hydrogen{} Infrastructure'' combine transport and storage for each carrier. ``Carbon Capture at Point Sources'' combines all technologies with integrated carbon capture, including the cost of the main facility (e.g., CHP unit) and the carbon capture application.}
    \label{fig:cost_bar}
\end{figure}

% In every model, nearly half the costs are attributable to wind and solar electricity production, and about 10\% to hydro, nuclear, and biomass. In the \baselinemodel, transmission costs (electricity, gas, hydrogen, carbon) are below 4\% (25 bn€) and are dominated by electricity grid expenses. In contrast, the other models have higher combined pipeline costs, peaking at 20\% in the \hybridmodel. Both carbonaceous fuel production and electrolysis account for about 4\% (30 bn€) and 4.7\% (37 bn€) of costs respectively, with heat pump installations contributing around 8\% (64 bn€ in the \baselinemodel).

System cost differences between the models are driven by wind and solar as well as carbon capture technologies. In scenarios with \carbon{} and \hydrogen{} transportation, the model primarily reduces the reliance on high-cost DAC from 392~Mt/a carbon removal in the \baselinescenario{} to 112~Mt/a in the \hybridscenario{} (see Fig.~\ref{fig:balance_captured_carbon}). At the same time, carbon capture from biomass combustion, gas-based industry processes and biogas-to-gas upgrading increase. The increased usage of upgraded biogas leads to a lower synthetic methane production.
% Another knock-on effect of less reliance on DAC is its reduced endogenous demand for dedicated renewables as well as heat pumps and biomass CHPs to supply power and heat.

\begin{figure}[ht!]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/balance_bar_carbon.png}
    \caption{Balance of captured carbon for all models in the Net-Zero scenario. Positive values indicate carbon capture, negative values indicate carbon consumption. By integrating \hydrogen{} and \carbon{} networks, the predominant method for carbon removal shifts from Direct Air Capture (DAC) to bio-energetic processes with capture. At the same time the reliance on methanation decreases.}
    \label{fig:balance_captured_carbon}
\end{figure}

In the \baselinescenario{}, where there is no carbon and hydrogen transport, hydrogen must be produced where it is consumed. For CU, either \hydrogen{} must be produced where \carbon{} is cheap at point sources, or DAC must be used at sites with low-cost \hydrogen{}. Where \carbon{} from point sources is not able to be sequestered, it has to be compensated elsewhere by negative emissions from DAC or biomass capture with sequestration. All of these options are sub-optimal.

@@

Both a carbon network and a hydrogen network can help alleviate these inefficiencies. A carbon network allows \carbon{} from point sources to be transported to low-cost hydrogen as well as sequestration sites. A hydrogen network on the other hand allows low-cost hydrogen to be transported from regions with good renewable resources to point sources for CU. This results in fundamentally different network flows in each scenario (see Fig.~\ref{fig:balance_map}).

\begin{figure*}[ht!]
    \centering
    \includegraphics[width=0.98\linewidth]{comparison/single-technologies/figures/90_nodes/balance_map_dedicated.png}
    \caption{Average production, consumption, flows and prices of the carbon (top line) and hydrogen (bottom line) sectors in the \carbongrid{} (left) and the \hydrogenscenario{} (right). For each region, upper semicircles show the average production per technology, lower semicircles the consumption, and colors the average marginal prices. Lines and arrows show the interregional transportation. Carbon sequestration offshore are drawn as full circles.
    }
    \label{fig:balance_map}
\end{figure*}

In the \carbonscenario, there are three major transport routes of the low-cost carbon captured in Central and Eastern Europe (see Fig.~\ref{fig:balance_map}, top left). A large part of the carbon is transported from Western Europe to the Iberian Peninsula (8~kt/h), the British Isles (7.5~kt/h) and Denmark (2.8~kt/h), where it is used along with low-cost electrolytic hydrogen to produce Fischer-Tropsch fuels and methanol. A second part is transported to methanation plants in Italy and Greece (2.8~kt/h), where the synthetic methane produced is fed into the gas grid in the vicinity of major gas import hubs. Finally, a third part of captured carbon is transported directly to sites in the North Sea, Baltic Sea and Mediterranean Sea where it is sequestered. Almost none of the captured carbon is used locally at the point-source. In addition to electrolysis in regions with the good renewable resources to supply CU ($\sim$125~GW$_\text{el}$), the model places smaller amounts of electrolyzers where hydrogen is needed for industrial processes across Central Europe ($\sim$8.5~GW$_\text{el}$) (see bottom left). Hydrogen prices vary strongly across the continent, with low price ``valleys'' where the CU locates with around 60~€/MWh, and high price ``peaks'' in Central Europe with around 95~€/MWh. Despite the transport system, carbon prices still vary from 110~€/t in Central Europe to 160~€/t at sequestration sites.

In the \hydrogenscenario{}, the main hydrogen transportation routes go from Spain (50~GW) and the United Kingdom (18~GW) to Central Europe to supply local carbonaceous fuel production with captured carbon from point sources (see Fig~\ref{fig:balance_map}, bottom right). Also, the \hydrogen{} network supplies spatially-fixed \hydrogen{} demand for industry in Central Europe (2.5~GW in total) at much lower costs than local electrolysis or SMR. Carbon sequestration primarily locates in Portugal where large DAC facilities operate (14~kt/h, see Fig~\ref{fig:balance_map}, top right). \carbon{} prices are low in Central Europe (110-120~€/t) and highest at sequestration regions (160~€/t). Hydrogen prices have flattened out at a relatively low level of €60~/MWh in all regions.

In both cases, CU is placed where material inputs, \carbon{} and \hydrogen{}, are provided at minimal cost, i.e., in price valleys of hydrogen and carbon in the \carbongrid{} and \hydrogenscenario{} respectively. In the \carbonscenario{}, CU co-locates with low-cost hydrogen production and uses transported carbon from the inland. In the \hydrogenscenario{}, CU co-locates with carbon point sources and uses transported low-cost hydrogen.
Both scenarios effectively combine low-cost renewable hydrogen production and carbon management for combustion and spatially-fixed industrial processes.

However, both transport systems provide individual cost-benefits to the system. While the carbon grid makes it possible to transport captured carbon for sequestration at low cost, the dependence on DAC near the coast causes higher costs in the hydrogen scenario{}. On the other hand, the \hydrogengrid{} is able to supply the spatially-fixed hydrogen demands in Central Europe with low-cost hydrogen imports, while the model in the \carbonscenario{} is forced to deploy electrolysis in regions with poor renewable resources as well as additional gas infrastructure in Central Europe. The latter aspect dominates the first, leading to lower system cost in the \hydrogenscenario{}.

\subsection*{Hybrid configuration provides further flexibilities}\label{subsec:Hybrid}

>>>>

\title{\hydrogen{} and \carbon{} Network Strategies for the European Energy System}
% \title{Competition and Synergies of \hydrogen{} and \carbon{} Networks in Europe}

@@

\begin{abstract}
    Hydrogen and carbon dioxide transport can both play an essential role in climate-neutral energy systems. Hydrogen networks help serve regions with high energy demand, while excess emissions are transported away in carbon dioxide networks. For the synthesis of carbonaceous fuels, it is less clear which input should be transported: hydrogen to carbon point sources or carbon to low-cost hydrogen. We explore both networks' potential synergies and competition in a cost-optimal carbon-neutral European energy system. In direct comparison, a hydrogen network is more cost-effective than a carbon network, since it serves to transport hydrogen to demand and to point source of carbon for utilization. However, in a hybrid scenario where both networks are present, the carbon network effectively complements the hydrogen network, promoting carbon capture from biomass and reducing reliance on direct air capture. Our analysis suggests integrating hydrogen and carbon dioxide networks into European energy policy for a robust, carbon-neutral or carbon-negative future energy system.
\end{abstract}

@@

The transition to a climate-neutral European economy is a pressing challenge that demands coordinated action across various energy sectors. While management of both carbon dioxide (\carbon{}) and hydrogen (\hydrogen{}) is considered a critical component of this transition, a gap exists in understanding how new hydrogen infrastructure effectively interacts with carbon management technologies, including carbon capture, transport, storage, utilization, and sequestration. Hydrogen is being considered in several industries that are not eligible for electrification. It offers an efficient way to transport and store energy over long distances and periods, and can be produced from electrolysis or steam methane reforming (SMR), for example. Captured carbon can be stored in geological formations to extract it from the atmosphere permanently, a process known as carbon sequestration (CS). Additionally, carbon can be combined with hydrogen to produce carbonaceous fuels, such as synthetic kerosene for aviation, synthetic methanol for shipping or synthetic methane for industrial feedstock. We jointly refer to the production of these carbonaceous fuels as Carbon Utilization (CU). Carbon can be effectively captured from industrial processes and the combustion of biomass, fossil fuels, or synthetic carbonaceous fuels through point-source carbon capture (CC) techniques, or harvested from the atmosphere using direct air capture (DAC). Both \carbon{} and \hydrogen{} networks are likely to play a crucial role in the cost-effective integration of these technologies to enable net-zero economies.

Recently, policymakers in Europe have started developing hydrogen and carbon management strategies~\cite{GermanyDevelopingStrategy2023,CarbonManagementStrategie}, planning infrastructure components~\cite{CONetz}, and committing to the first carbon utilization projects~\cite{EFuelsPilotPlant2022,OrstedAssumesFull,GROUNDBREAKINGEFUELPRODUCTION,DLREfuelsDLR}. With the European Union's goal of achieving climate neutrality by 2050, a wide range of programs, funding models and initiatives have been established with dedicated instruments supporting both the ramp-up of the hydrogen and carbon economy~\cite{eu2023netzero,europeangreendeal,europeaninnovationfund}.
% TODO: split into hydrogen and carbon?
% What about NZIA "proposes a CO2 injection target of 50 million tonnes per year by 2030, which will be stored within the European Union (EU)"
% What about RED III targets, 10 MtH2/a domestic in 2030? Name a few targets rather than hiding everything in references - readers won't go through all references..

Collaborative industry initiatives like the European Hydrogen Backbone~\cite{gasforclimateEuropeanHydrogenBackbone2022} or the Hydrogen Infrastructure Map~\cite{H2InfrastructureMap} showcase the potential of hydrogen as a fuel and energy carrier. Some natural gas pipelines have already been repurposed to transport hydrogen~\cite{RohrFreiFuer}. Business models from companies like Tree Energy Solutions~\cite{TESHydrogenLife2023}, Carbfix~\cite{WeTurnCO2}, and Equinor~\cite{adomaitisEquinorRWEBuild2023} advertise carbon management hubs that provide green hydrogen, methane, and synthetic fuels on the one hand and offer to purchase \carbon{} on the other hand. The Northern Lights project in Norway~\cite{NorthernLightsWhat} is planning with a transport and sequestration capacity of 1.5~Mt \carbon{} per year to be operative in 2024, expanding to a targeted scale of 5~Mt per year of sequestration by 2030.
The market potential for carbon capture and sequestration in Europe is underscored by the Capture Map's estimated potential~\cite{ToolsGreenTransition} of 1.7 Gt carbon annually from point sources, accounting for about half of the continent's emissions, as well as estimated sequestration capacity potentials of up to 3~Gt/a~\cite{europeancommissionEuropeanCO2Storage}. However, it's important to note that carbon sequestration technology, unlike carbon capture, is not yet fully mature and its full potential and implications remain to be explored in depth.
To advance the carbon economy, the Clean Air Task Force emphasizes the importance of developing a carbon transport system in Europe alongside a hydrogen network~\cite{lockwoodEuropeanStrategyCarbon}. Carbon pipelines, considered a mature technology, have seen widespread installations in the United States and Canada, primarily to supply \carbon{} to enhanced oil recovery~\cite{righettiSitingCarbonDioxide2017,friedmannNETZEROGEOSPHERICRETURN}.

Up to this point, it remains unclear how the two transport systems of hydrogen and carbon may complement or even compete with each other. Both can bridge the misalignment of sources and sinks for carbon and hydrogen. A hydrogen network can supply regions with geographically fixed hydrogen demand, such as for steel production, with hydrogen from regions with the best renewable resources as well as enable CU at the site of point-source CC. On the other hand, the carbon transport system allows transporting captured carbon to regions with sequestration potentials or high-quality renewable resources, the latter enabling cost-effective CU.

In the literature, the two network approaches and underlying technologies have been discussed in several publications, all of which, however, dealt with the isolated aspects~\cite{bakkenLinearModelsOptimization2008,morbeeOptimisedDeploymentEuropean2012,stewartFeasibilityEuropeanwideIntegrated2014,oeiModelingCarbonCapture2014,elahiMultiperiodLeastCost2014,burandtDecarbonizingChinaEnergy2019,middletonSimCCSOpensourceTool2020,bjerketvedtOptimalDesignCost2020,weiProposedGlobalLayout2021,damoreOptimalDesignEuropean2021,becattiniCarbonDioxideCapture2022,neumannBenefitsHydrogenNetwork2022}. Such techno-economic models, in contrast for example to Integrated Assessment Models, can account for the spatial distribution of carbon sources and sinks, which are crucial to provide a holistic view of the energy system and its technological interactions. Neumann et al.~\cite{neumannBenefitsHydrogenNetwork2022} demonstrated the interchangeability of hydrogen and electricity grid expansion in a climate-neutral European energy system at the expense of higher investments. The underlying, highly-resolved model encompasses the most relevant sectors, however it neglects the option to transport carbon across regions.
Morbee et al.~\cite{morbeeOptimisedDeploymentEuropean2012} optimizes the topology and capacity of a \carbon{} network in Europe, but only considers the power sector without co-optimizing renewable deployment. This limited sectoral scope cannot capture important dynamics of carbon management, since it neglects sectors like industry that will need to handle most \carbon{} in the future.
A comprehensive example is found in~\cite{becattiniCarbonDioxideCapture2022}, which presents a mixed-integer model to optimize the time-evolution of a \carbon{} transport system in Switzerland, connected to a remote sequestration site in Norway. However, this limited spatial scope fails to consider other sequestration sites and the co-benefits from connecting the \carbon{} network to neighboring countries.
Other high-resolution energy system models do not feature detailed CU technologies and \carbon{} transport. Euro-Calliope, for example, does not feature carbon nor hydrogen transport, and thus forces generic synthetic fuel production to rely on captive electrolysis and DAC in the same region~\cite{pickeringDiversityOptionsEliminate2022}.

No study has yet considered the co-optimization and comprehensive assessment of both \carbon{} and \hydrogen{} networks in a fully sector-coupled energy system. However, such an assessment is strongly needed to identify realms of competition and synergy between hydrogen and carbon management technologies. In this paper, we present a detailed study of the European energy system for 2050, which includes high geographical resolution and a comprehensive representation of carbon management technologies. The study is conducted using the PyPSA-Eur energy system model~\cite{brownPyPSAEurOpenSectorcoupled2023} which optimizes operations and investments in all relevant energy sectors to supply projected energy demands for 2050 (see Fig.~\ref{fig:total-demand-bar}). It is assumed that Europe is self-sufficient in energy and does not import any fuels. This drives carbonaceous fuel production to be located at sites where both \carbon{} and \hydrogen{} can be provided at low cost.

\begin{figure}[h!]
    \includegraphics[width=\linewidth]{baseline/figures/90_nodes/demand_bar.png}
    \caption{Assumptions on exogenous demand, derived from~\cite{piamanzGeoreferencedIndustrialSites2018,muehlenpfordtTimeSeries2019,mantzosJRCIDEES20152018,NationalEmissionsReported2023,EurostatCompleteEnergyBalance,uwekrienDemandlib2023}. The figure shows total annual energy demands for each energy source, which determine the model's endogenous investments and operation. Endogenous processes can lead to higher total production volumes of some energy carriers, e.g., the demand for methanol requires more hydrogen and carbon as secondary (energy) inputs, which are not considered here. In the model, demands are defined per region and time stamp.}
    % TODO: adjust labels to show that these are exogenous assumptions
    \label{fig:total-demand-bar}
\end{figure}

@@

To investigate the competing transport dynamics between \carbon{} and \hydrogen{} networks, we contrast the following four scenarios:
%
\begin{itemize}
    \item[] \textit{Baseline:} Neither \carbon{} nor \hydrogen{} networks are constructed.
    \item[] \textit{\carbongrid{}:} Only the \carbon{} network is developed.
    \item[] \textit{\hydrogengrid{}:} Only the \hydrogen{} network is developed.
    \item[] \textit{Hybrid:} Both \carbon{} and \hydrogen{} networks are developed.
\end{itemize}

We analyze how the deployment of carbon and hydrogen networks affects different decarbonization strategies in the scenarios and how investments in none or only one of the transport technologies lead to inefficient market layouts. We do this by focusing first on a net-zero \carbon{} emissions target, while limiting carbon sequestration to 200~Mt/a. We give a detailed description of the different technological and geographical impacts of the transportation systems. Subsequently, our analysis demonstrates the extent to which different deployment strategies remain robust under a tightened net \carbon{} removal target.

@@

All scenarios share key energy system characteristics but differ in system costs and technology use (Fig.~\ref{fig:cost_bar}). Costs in the \baselinescenario{} are the highest at 779~bn€/a. Scenarios with carbon and hydrogen networks are more cost-effective, reducing costs by 3.1\% in the \carbongrid{}, 4.8\% in the \hydrogengrid{}, and 5.2\% in the \hybridscenario{}.

\begin{figure}[ht!]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/cost_bar.png}
    \caption[short]{Total annual system cost, subdivided into groups of technologies for the different models of the European energy system with a net-zero emission target. While in the \baselinescenario{}, the model has neither a carbon nor a hydrogen network, it can expand both in the \hybridscenario{}. ``Gas Infrastructure'' combines gas facilities for power and heat production, ``\carbon{} Infrastructure'' and ``\hydrogen{} Infrastructure'' combine transport and storage for each carrier. ``Carbon Capture at Point Sources'' combines all technologies with integrated carbon capture, including the cost of the main facility (e.g., CHP unit) and the carbon capture application.}
    \label{fig:cost_bar}
\end{figure}

% In every model, nearly half the costs are attributable to wind and solar electricity production, and about 10\% to hydro, nuclear, and biomass. In the \baselinemodel, transmission costs (electricity, gas, hydrogen, carbon) are below 4\% (25 bn€) and are dominated by electricity grid expenses. In contrast, the other models have higher combined pipeline costs, peaking at 20\% in the \hybridmodel. Both carbonaceous fuel production and electrolysis account for about 4\% (30 bn€) and 4.7\% (37 bn€) of costs respectively, with heat pump installations contributing around 8\% (64 bn€/a in the \baselinemodel).

System cost differences between the scenarios are driven by wind and solar, as well as carbon capture technologies. In scenarios with carbon and hydrogen networks, the model primarily reduces the reliance on high-cost DAC, from 392~Mt/a carbon removal in the \baselinescenario{} to 112~Mt/a in the \hybridscenario{} (see Fig.~\ref{fig:balance_captured_carbon}). At the same time, carbon capture from biomass combustion, gas-based industry processes and biogas-to-gas upgrading increase, the latter lowering the need for synthetic methane production.
% Another knock-on effect of less reliance on DAC is its reduced endogenous demand for dedicated renewables as well as heat pumps and biomass CHPs to supply power and heat.

\begin{figure}[ht!]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/balance_bar_carbon.png}
    \caption{Balance of captured carbon for all scenarios assuming net \carbon{} neutrality. Positive values indicate carbon capture, negative values indicate carbon consumption. By integrating hydrogen and carbon networks, the predominant method for carbon removal shifts from Direct Air Capture (DAC) to bio-energetic processes with capture. At the same time, the reliance on methanation decreases.}
    \label{fig:balance_captured_carbon}
\end{figure}

In the \baselinescenario{}, where there is no carbon and hydrogen transport, hydrogen must be produced where it is consumed. For CU, either hydrogen must be produced where carbon is cheap at point sources, or DAC must be used at sites with low-cost hydrogen. Where carbon from point sources cannot be sequestered, it has to be compensated elsewhere by negative emissions from DAC or biomass capture with sequestration. All of these options are sub-optimal from a cost-perspective.

@@

Both a carbon network and a hydrogen network can help alleviate these inefficiencies. On the one hand, a carbon network allows carbon from point sources to be transported to low-cost hydrogen as well as to sequestration sites. On the other hand, a hydrogen network allows low-cost hydrogen to be transported from regions with good renewable resources to point sources for CU. This results in fundamentally different network flows in each scenario (see Fig.~\ref{fig:balance_map}).

\begin{figure*}[ht!]
    \centering
    \includegraphics[width=\linewidth]{comparison/single-technologies/figures/90_nodes/balance_map_dedicated.png}
    \caption{Average production, consumption, flows and prices of the carbon (top line) and hydrogen (bottom line) sectors in the \carbongrid{} (left) and the \hydrogenscenario{} (right). For each region, upper semicircles show the average production per technology, lower semicircles the consumption, and colors the average marginal prices. Lines and arrows show the interregional transportation. Carbon sequestration offshore is drawn in full circles.
    }
    \label{fig:balance_map}
\end{figure*}

In the \carbonscenario, there are three main transport purposes of the low-cost carbon captured in Central and Eastern Europe (see Fig.~\ref{fig:balance_map}, top left). A large part of the carbon is transported to the Iberian Peninsula (8~kt/h), the British Isles (7.5~kt/h) and Denmark (2.8~kt/h), where it is used along with low-cost electrolytic hydrogen to produce Fischer-Tropsch fuels and methanol. A second part is transported to methanation plants in Italy and Greece (2.8~kt/h), where the synthetic methane is fed into the gas grid near large, geographically defined terminals where fossil gas is imported from outside Europe. Finally, a third part of captured carbon is transported directly to sites in the North Sea, Baltic Sea and Mediterranean Sea where it is sequestered. Almost none of the captured carbon is used locally at the point-source. In addition to electrolysis in regions with good renewable resources to supply CU ($\sim$125~GW$_\text{el}$), the model places smaller amounts of electrolyzers where hydrogen is needed for industrial processes across Central Europe ($\sim$8.5~GW$_\text{el}$) (see bottom left). Hydrogen prices vary widely across the continent, with low price ``valleys'' at around 60~€/MWh, where the CU is located, and high price ``peaks'' in Central Europe at around 95~€/MWh. Despite the transport system, carbon prices still vary from 110~€/t in Central Europe to 160~€/t at sequestration sites.

In the \hydrogenscenario{}, the main hydrogen transportation routes go from Spain (50~GW) and the United Kingdom (18~GW) to Central Europe to supply local carbonaceous fuel production with captured carbon from point sources (see Fig~\ref{fig:balance_map}, bottom right). Also, the hydrogen network supplies spatially-fixed hydrogen demand for industry in Central Europe (2.5~GW in total) at much lower costs than local electrolysis or SMR. Carbon sequestration is primarily located in Portugal where large DAC facilities operate (14~kt/h, see Fig~\ref{fig:balance_map}, top right). Carbon prices are low in Central Europe (110-120~€/t) and highest in sequestration regions (160~€/t). Hydrogen prices have flattened out at a relatively low level of €60~/MWh in all regions.

In both scenarios, CU is placed where material inputs, \carbon{} and \hydrogen{}, are provided at minimal cost, i.e., in price valleys of hydrogen and carbon in the \carbongrid{} and \hydrogenscenario{} respectively. In the \carbonscenario{}, CU co-locates with low-cost hydrogen production and uses transported carbon from the inland. In the \hydrogenscenario{}, CU co-locates with carbon point sources and uses transported low-cost hydrogen.
Thus, both scenarios effectively combine low-cost renewable hydrogen production and carbon management for combustion and spatially-fixed industrial processes.

However, there are individual cost-benefits. On the one hand, the carbon network enables transport of low-cost captured carbon to sequestration sites, reducing the dependence on high-cost DAC near the coast which are present in the \hydrogenscenario{}. On the other hand, the \hydrogengrid{} is able to supply the spatially-fixed hydrogen demands in Central Europe with low-cost hydrogen imports, while the model in the \carbonscenario{} is forced to deploy electrolysis in regions with poor renewable resources as well as additional gas infrastructure in Central Europe. The latter aspect finally leads to lower system costs in the \hydrogenscenario{} than in the \carbonscenario{}.

\subsection*{Hybrid configuration combines advantages}\label{subsec:Hybrid}

\begin{figure*}[ht!]
    \centering
    \begin{subfigure}{.5\textwidth}
        \centering
        \includegraphics[width=\linewidth]{full/figures/90_nodes/balance_map_carbon.png}
        \label{fig:balance_map_carbon_full}
    \end{subfigure}%
    \begin{subfigure}{.5\textwidth}
        \centering
        \includegraphics[width=\linewidth]{full/figures/90_nodes/balance_map_hydrogen.png}
        \label{fig:balance_map_hydrogen_full}
    \end{subfigure}
    \caption{Average operation, flows and prices of the carbon (left) and hydrogen (right) sectors in the \hybridscenario{} assuming net \carbon{} neutrality. For each region, upper semicircles show the average production per technology, lower semicircles the consumption, and colors the average marginal prices. Carbon Sequestration offshore is drawn in full circles. Lines and arrows show interregional transportation. Carbon from point-source in the inland either supplies local CU with imported hydrogen or facilitates sequestration in nearby offshore regions.
    % Carbon network looks the same as in~\cite{morbeeOptimisedDeploymentEuropean2012}: two backbones, one in the nothern Europe other in south east.
    }
    \label{fig:balance_map_full}
\end{figure*}
